---
title: "Multivariate Analysis"
author:
  - Ana Rasinhas
  - Daniela Gonçalves
  - Pedro Fernandes
  - Ricardo Branco
  - Samuel Medalha
date: last-modified
format:
  pdf: default
execute:
  echo: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(readxl)
library(psych)
```

```{r}
#| label: import-data
#| include: false

selected_vars <- c("cntry", "trstprl", "trstlgl", "trstplc", "trstplt", "trstprt", "trstep", "trstun", "stflife", "stfeco", "stfgov", "stfdem", "stfedu", "stfhlth", "imbgeco", "imueclt", "imwbcnt")
country_data <- na.omit(read_excel("./dataset1.xlsx", sheet = 1, na = c("77", "88", "99"))[, selected_vars]) %>% 
  rename(
    X1 = trstprl, X2 = trstlgl, X3 = trstplc, X4 = trstplt, X5 = trstprt, X6 = trstep, X7 = trstun, X8 = stflife, X9 = stfeco, X10 = stfgov, X11 = stfdem, X12 = stfedu, X13 = stfhlth, X14 = imbgeco, X15 = imueclt, X16 = imwbcnt
  )
data <- country_data %>% select(-cntry)
```


## Abstract

## Introduction

## Methodology
```{r}
#| label: KMO
#| warning: false
KMO(data)
```
After executing the Kaiser-Meyer-Olkin test, the MSA (measure of sampling accuracy, which varies between 0 and 1) is calculated as `r format(KMO(data)$MSA, digits = 2)`. This is classified as "meritorious", so we have an indication that factor analysis is suitable for this data set.

We'll start by performing factor analysis with a high number of values:

```{r}
#| label: first_fa
#| warnings: false
#| echo: true
fa <- fa(data, nfactors = 8, rotate = "varimax") 
print(fa$e.values, sort = TRUE, decreasing = TRUE)
```
There are 3 eigenvalues greater than 1.

```{r}
#| label: cumulative_var
print(fa$Vaccounted, digits = 2)
```
However, 10 factors aren't enough to explain 80% of the variance (???)

```{r}
#| label: scree_plot
scree(data, factors = FALSE)
```

```{r}
fa_3_varimax <- fa(data,
  nfactors = 6,
  fm="pa",
  rotate = "varimax")
```

```{r}
fa.diagram(fa, errors = FALSE)
```

```{r}
factor_scores <- factor.scores(f = fa_3_varimax, x = data)
scores <- cbind(country_data, factor_scores$scores) %>% group_by(cntry) %>% summarize(PA1 = mean(PA1), PA2 = mean(PA2), PA3 = mean(PA3), PA4 = mean(PA4), PA5 = mean(PA5), PA6 = mean(PA6))
```




## Results

## Discussion

## Conclusions

## References